Individual Variation in Animal Movement

### 2021-03-31 distributions_paper/rudolph_distributions.Rmd
finished draft of introduction

# Introduction
   Movement is a fundamental ecological process, critical to an animal's biology and it's interactions with the environment. The study of animal movement tends to focus on four basic mechanistic questions: the why, when, where, and how an animal moves, and these questions can be answered at the scale of individuals or populations. A general framework for movement ecology is needed to link movement patterns and processes. Such a framework would start with a focus on movement itself and the individual's internal state, navigation capacity, motion capacity, and responses to external factors [@nathan2008movement]. The overall goal of studying movement is to find the linkages between the causes and mechanisms of animal movement, that explain spatiotemporal patterns that then feed back into various ecological and evolutionary processes. Some of these approaches involve separating different movement patterns into behavioral states, with each state having an associated set of parameters that determine that specific movement pattern[@langrock2012flexible]. Other approaches have focused on segmenting movement patterns, again in an attempt to associate specific types of movement to underlying behaviors or fine-scale environmental influences (Getz citations, Morales, more focus on the random walk models, Patterson2008). *Indeed, animal movement has consequences.  Such consequences might influence species interactions, ecological function, and vital demographic processes (survival, reproduction, dispersal) that can result in population impacts and evolutionary change (?).*

  Although there is variation in animal movement at the individual scale, the general focus has been on understanding the typical movement patterns, built around average measures at population levels, with the simplest approach assuming model parameters are the same across individuals, known as 'complete pooling' [@langrock2012flexible]. This focus on generality has driven to *a lack of frameworks* for understanding variation, how it is maintained, and the consequences of this variation across populations, communities, and ecosystems. Some alternatives have been considered to incorporate individual variation with 'no pooling' or 'partial pooling' approaches, where each or some individuals have their own set of parameters [@patterson2009classifying, @jonsen2006robust], however a high number of parameters quickly limits applications to high levels of individual variation. Recent calls for incorporating individual variation in movement [@snell2019consequences, @shaw2020causes] have pushed for the need to develop theory that explores different types of variation in movement patterns to understand its consequences across ecological organizational levels. In particular, when considering species interactions, incorporating individual variation becomes essential to understand how positive or negative feedback loops play into the maintenance of variation in movement as individuals respond to external factors in their environment [@shaw2020causes].

@@

# Kareiva Shigesada paper [@kareiva1983analyzing]

We focus here on developing a process to analyze animal movement sequences based on movement lengths, or step lengths, and turning angle probability distributions. Following the methods of Kareiva and Shigesada [-@kareiva1983analyzing], we assume the movement is a correlated random walk and derive expected square displacements associated to a number of consecutive moves under various probability distribution models. Unlike other studies which focus on deciphering the underlying behaviors from animal movement data, our focus was to uncover potential differences between individuals and how these differences can be obscured when modeling population-level animal movement.

@@

"We know these simple correlated random walks, or simple movement models might not be enough to describe net displacements under more complex processes, for which developments in higher order markov processes show to be promising. However, our goal in this study was to use a simpler model and focus on potential differences when analyzing complete pooling versus no pooling data. In the context of animal-mediated seed dispersal, understanding these indivdiual level differences can help us describe and characterize the implications that different animals may have over spatial distributions of plant populations, and thus identify potential dispersers with higher or more importance, particularly when associated to long-distance seed dispersal events." Written by Javi

@@

A simple correlated random walk, which involves a directional correlation between successive step movement (coddling2008)

@@

@@

aracari_df %>% 
  mutate(tmin = dt/60,
         mpm = dist/tmin) %>% 
  drop_na(tmin) %>% 
  filter(tmin == 15) %>% 
  group_by(id) %>% 
  tally() %>% 
  filter(n >=20)

>>>>

# Introduction
   Movement is a fundamental ecological process, critical to an animal's biology and it's interactions with the environment. The study of animal movement tends to focus on four basic mechanistic questions: the why, when, where, and how an animal moves, and these questions can be answered at the scale of individuals or populations. A general framework for movement ecology is needed to link movement patterns and processes. Such a framework would start with a focus on movement itself and the individual's internal state, navigation capacity, motion capacity, and responses to external factors [@nathan2008movement]. The overall goal of studying movement is to find the linkages between the causes and mechanisms of animal movement, that explain spatiotemporal patterns that then feed back into various ecological and evolutionary processes. Some of these approaches involve separating different movement patterns into states, with each state having an associated set of parameters that determine that specific movement pattern [@langrock2012flexible]. These approaches segment movement patterns, in order to associate specific types of movement to underlying behaviors or fine-scale environmental influences (Getz citations, Morales, more focus on the random walk models, Patterson2008). Several of these approaches focus on the why and underlying processes that drive animal movement, whether that is an animal's internal state or responses to their environment.

   Random walks have been used broadly in biology to characterize animal movement, and increasing levels of complexity can be incorporated by using multiple random walks, continuous time or different levels of correlation between turning angles [@codling2008random; @morales2004extracting]. More recent developments have focused on implementing higher order Markov processes through state-space models, which provide a probabilistic framework that allows for prediction of future states of a system based from previous states, by coupling an observation model with an underlying mechanistic model. When extending this framework to movement ecology, we understand that the process model provides the mechanism for movement across space and time, and the observation model is related to the sampling process. In movement data, a state-space process allows us to couple a statistical model for the observation method, accounting for errors in sampling or detection, with a separate model for the movement dynamics which are determined by the effects of behavior or responses to the environment [@patterson2008state]. In general, state-space models are used to identify different states, or behavioral states, that determine different movement patterns associated to step length and turning angle. And some of the main questions asked with this approach are related to identifying these different states, or behavioral state-switching, from time series data on location tracking.

   Although I have mostly described approaches used to understand why movement happens, animal movement also has consequences, which might influence species interactions, ecological function, and vital demographic processes (such as survival, reproduction, or dispersal of other species) and thus result in population or regional level impacts. A clear example of the consequences of animal movement on other species involves animal-mediated seed dispersal or pollination, impacting a plant's spatial population dynamics.The specific case of seed dispersal is interesting because of the different levels and interactions happening. At the individual level, we focus on an individual bird consuming fruit and dispersing the seeds of the tree. How this particular individual moves across the landscape directly affects the dispersal for that specific tree and other trees from which it consumes fruit. At the same time, that individual has intra-individual variation in movement patterns associated to specific events in that individual's lifetime, such as breeding and nesting, or even smaller scale variation related to different states, such as foraging or resting. Interspecific interactions come at play as multiple species find themselves together feeding from the same trees, and thus competing for these resources, while at the same time intraspecific interactions due to social behavior broadly determine general movement patterns for animals restricted within family home ranges. From the plant's perspective, the variation in movement patterns across all dispersers will determine eventual plant population spatial spread.

  Specifically in the context of seed dispersal ecology, the general focus has been on understanding how the typical movement patterns, built around average measures at population levels, can influence plant population spread and seed dispersal [sources]. Animal movement can be modeled with a simple diffusion process or a random walk [@jones2017closing; @levey2005effects], with the simplest approach assuming model parameters are the same across individuals, something known as 'complete pooling' [@langrock2012flexible]. This focus on generality has driven to a lack of frameworks for understanding how variation in animal movement is maintained, and the consequences of this variation across populations and communities. From the animal movement perspective, some alternatives have been considered to incorporate individual variation with 'no pooling' or 'partial pooling' approaches, where each or some individuals have their own set of parameters [@patterson2009classifying, @jonsen2006robust], however a high number of parameters quickly limits applications to high levels of individual variation. Recent calls for incorporating individual variation in movement and their consequences to seed dispersal or even pathogen spread [@snell2019consequences, @shaw2020causes] have pushed for the need to develop theory that explores different types of variation in movement patterns to understand its consequences across ecological organizational levels. In particular, when considering species interactions, incorporating individual variation becomes essential to understand how positive or negative feedback loops play into the maintenance of variation in movement as individuals respond to external factors in their environment [@shaw2020causes].

  We focus here on developing a process to analyze animal movement sequences based on movement lengths, or step lengths, and turning angle probability distributions. Following the methods of Kareiva and Shigesada [-@kareiva1983analyzing], we assume the movement is a correlated random walk and derive expected square displacements associated to a number of consecutive moves under various probability distribution models.A simple correlated random walk, which involves a directional correlation between successive step movement [@codling2008random]. Unlike other studies which focus on deciphering the underlying behaviors from animal movement data, our focus was to uncover potential differences between individuals and how these differences can be obscured when modeling population-level animal movement. We know these simple correlated random walks, or simple movement models might not be enough to describe net displacements under more complex processes, for which developments in higher order markov processes show to be promising. However, our goal in this study was to use a simpler model and focus on potential differences when analyzing complete pooling versus no pooling data. In the context of animal-mediated seed dispersal, understanding these indivdiual level differences can help us describe and characterize the implications that different animals may have over spatial distributions of plant populations, and thus identify potential dispersers with higher or more importance, particularly when associated to long-distance seed dispersal events.

@@

@@

@@

@@

$$AIC{_{indlevel}} = 2 \cdot \sum \left ( number parameters \right ) + 2 \cdot \sum \left ( negative log likelihood \right )$$

$$ AICc{_{indlevel}} =  AIC{_{indlevel}} + (2 \cdot \sum{nparams}  \cdot \left(\sum{nparams}+1\right)/\left(\sum{observations} - \left(\sum{nparams} - 1\right)\right)$$

$$BIC_{indlevel} = \ln{\sum observations} \cdot \sum{number parameters} + 2\cdot \sum {negative log likelihood}$$

@@

aracari_df %>% 
  mutate(tmin = dt/60,
         mpm = dist/tmin) %>% 
  drop_na(tmin) %>% 
  filter(tmin == 60) %>% 
  group_by(id) %>% 
  tally() %>% 
  filter(n >=20)

### 2021-03-31 movement_paper/rudolph_ch2.Rmd
finished draft of introduction

## How does this relate to mutualisms and long-distance movements?
  The specific case of seed dispersal is interesting because of the different levels and interactions happening. At the individual level, we focus on an individual bird consuming fruit and dispersing the seeds of the tree. How this particular individual moves across the landscape directly affects the dispersal for that specific tree and other trees from which it consumes fruit. At the same time, that individual has intra-individual variation in movement patterns associated to specific events in that individual's lifetime, such as breeding and nesting, or even smaller scale variation related to different states, such as foraging or resting. Interspecific interactions come at play as multiple species find themselves together feeding from the same tress, and thus competing for these resources, while at the same time intraspecific interactions due to social behavior broadly determine general movement patterns. From the plant's perspective, the variation in movement patterns across all dispersers will determine eventual plant population spatial spread and success. Individuals will vary in their distances traveled while foraging or searching, directly influencing seed dispersal for the plant. Individuals with a tendency to travel larger distances or with higher probabilities of long-distance movements, will have a larger impact on plant population dynamics by increasing the plant's long-distance dispersal. These long-distance movements for seed-dispersing animals, or long-distance dispersal events for plants, can lead to spatial sorting of populations and range expansions [These are sources I haven't finished reading yet: Shine2011PNAS, Philips2008AmNat, Riotte-Lambert2019Trends.] This is also the case for pollination, where pollinator variability in movement patterns and long-distance movements has direct implications for pollination success and plant population persistence, specially in fragmented landscapes [need to find sources for this]. 
  Shaw makes references to how interspecific interactions can contribute to individual movement variation in parasite/pathogen systems. In my case, I would consider how presence of other competitive frugivores could influence this variation (Competition between toucans for the virola fruits and space. Larger aracari kick out the little ones [@holbrook2011home]). Shaw brings up how worse conditions can cause increased movement for searching better habitats and escape from these areas.

@@

## Overview of state-space models for movement  
  State-space models provide a probabilistic framework that allow for prediction of future states of a system based from previous states, by coupling an observation model with an underlying mechanistic model. When extending this framework to movement ecology, we understand that the process model provides the mechanism for movement across space and time, and the observation model is related to the sampling process. In movement data, a state-space process allows us to couple a statistical model for the observation method, accounting for errors in sampling or detection, with a separate model for the movement dynamics which are determined by the effects of behavior or responses to the environment [@patterson2008state]. In general, state-space models are used to identify different states, or behavioral states, that determine different movement patterns associated to step length and turning angle. And some of the main questions asked with this approach are related to identifying these different states, or behavioral state-switching, from time series data on location tracking.HMMs can be used to categorize this movement data into behavioral states, and link it to foraging or mating activity, as an example.

>>>>

## How does this relate to mutualisms and long-distance movements?
   Individuals will vary in their distances traveled while foraging or searching, directly influencing seed dispersal for the plant. Individuals with a tendency to travel larger distances or with higher probabilities of long-distance movements, will have a larger impact on plant population dynamics by increasing the plant's long-distance dispersal. These long-distance movements for seed-dispersing animals, or long-distance dispersal events for plants, can lead to spatial sorting of populations and range expansions [These are sources I haven't finished reading yet: Shine2011PNAS, Philips2008AmNat, Riotte-Lambert2019Trends.] This is also the case for pollination, where pollinator variability in movement patterns and long-distance movements has direct implications for pollination success and plant population persistence, specially in fragmented landscapes [need to find sources for this]. 
  Shaw makes references to how interspecific interactions can contribute to individual movement variation in parasite/pathogen systems. In my case, I would consider how presence of other competitive frugivores could influence this variation (Competition between toucans for the virola fruits and space. Larger aracari kick out the little ones [@holbrook2011home]). Shaw brings up how worse conditions can cause increased movement for searching better habitats and escape from these areas.

@@
